Methods paper for erosion

### 2017-07-14 docs/methods/erosion_methods.tex
cleaned up fortran code and added a routine to compute the velocity and pressure at target points

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation\label{s:formulation}} 
We consider multiple bodies being eroded by a two-dimensional Stokesian
fluid.  The Stokes paradox is circumvented by using a rounded-off
rectangle that encloses the eroding bodies.  Then, the flow is driven
by a boundary conditions, such as a Hagen-Poiseuille flow, on the
boundary of the outer geometry \todo[inline]{Need a figure here}
\begin{subequations}
\label{eqn:erosionModel}
\begin{gather}
eqn1 \label{eqn:eqn1} \\
eqn2 \label{eqn:eqn2} \\
eqn3 \label{eqn:eqn3}
\end{gather}
\end{subequations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
The incompressible Stokes equations are
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Gamma.
\end{alignat*}
The viscosity $\mu$ sets the time scale, and we will assume that
$\mu=1$.  The double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\Gamma} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:dlp}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the unit outward
normal, solves the incompressible Stokes equations for any sufficiently
smooth density function $\eeta$.  However, the double-layer potential can not capture velocity fields
corresponding to rigid body motions~\cite{pow-mir1987}.  A standard
method to complete the double-layer potential is to introduce a
Stokeslet and rotlet due to each interior body
\begin{align}
  S[\llambda_k,\cc_k] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_k
  \quad \text{and} \quad
  R[\xi_k,\cc_k] = \frac{\rr^\perp}{\rho^2}\xi_k,
  \label{eqn:stokeslet_rotlet}
\end{align}
where $\cc_k$ is a point in the interior of $\Omega_k$, $\rr = \xx -
\cc_k$, and $\rr^\perp = (r_2,-r_1)$.  Then, the solution of the Stokes
equations with a Dirichlet boundary condition $\ff$ can be written in
terms of the completed double-layer potential
\begin{align}
  \uu(\xx) = \DD[\eeta](\xx) + \sum_{k=1}^{M} S[\llambda_k,\cc_k](\xx) +
              \sum_{k=1}^{M} R[\xi_k,\cc_k](\xx), \quad \xx \in \Omega,
  \label{eqn:completedDLP}
\end{align}
where the density function, Stokeslets, and rotlets satisfy the
second-kind integral equation
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{k=1}^{M} S[\llambda_k,\cc_k](\xx) +
      \sum_{k=1}^{M} R[\xi_k,\cc_k](\xx), \quad \xx \in \Gamma, \\
  \llambda_k &= \frac{1}{2\pi}\int_{\Gamma_k} \eeta(\yy) \, ds_\yy, \\
  \xi_k &= \frac{1}{2\pi}\int_{\Gamma_k} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy.
  \label{eqn:completed_DLP}
\end{align}
Finally, because of the incompressibility constraint, the boundary
condition must satisfy 
\begin{align}
  \int_{\Gamma} \ff \cdot \nn \, ds_\yy = 0.
\end{align}
This results in a rank-one null space of
equation~\eqref{eqn:completed_DLP} which we remove by adding the rank
one modification $\NN_0[\eeta](\xx) = \int_{\Gamma_0} (\nn(\xx) \times
\nn(\yy))\nn(\yy) \, ds_\yy$, for $\xx \in \Gamma_0$.

@@

Considering the double-layer potential~\eqref{eqn:dlp} evaluated at $\xx
\in \Omega$, the deformation tensor is
\begin{align*}
  \ssigma^\DD(\xx) = \int_{\bd\Omega} K(\xx,\yy) \eeta(\yy) \,ds_\yy,
\end{align*}
where the kernel $K$ is given by
\begin{align*}
  K(\xx,\yy) = \frac{1}{2\pi}\left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn) +
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
We are interested in the deformation tensor at boundary points $\xx_0
\in \bd\Omega$, and this requires an expression for the jump in
$\ssigma^\DD(\xx)$ as $\xx \rightarrow \xx_0 \in \Gamma$.  The limiting
value is given by~\cite{qua-bir2014a}
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}\sigma_\DD(\xx) =
  J[\eeta](\xx_0) + \ssigma^\DD(\xx_0) = 
  \frac{1}{2}\pderiv{\eeta}{\ss} \cdot \ss
  \left[ \begin{array}{cc}
    s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
  \end{array}\right] + \ssigma^\DD(\xx_0), \quad \xx_0 \in \Gamma.
\end{align}

@@

To compute the pressure at $\xx_0 \in \Gamma$, we follow the same
procedure used for the shear stress.  In particular, we compute the
pressure due to each of the three terms that compose the completed
double-layer potential.  They are given by
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}p^\DD(\xx) =  
  -\pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0),
\end{align}
where
\begin{align}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\Gamma} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta \,ds_\yy.
\end{align}
The pressure due to the Stokeslets is
\begin{align}
  p^S(\xx_0) = \frac{\rr \cdot \eeta}{2\pi\rho^2},
\end{align}
and the velocity due to a rotlet is pressure-free.  Therefore, once the
density function and Stokeslets are computed, the pressure at a point
$\xx_0 \in \Gamma$ is
\begin{align}
  p(\xx_0) = -\pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0) + 
              \frac{\rr \cdot \eeta}{2\pi\rho^2}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
We discretize the second-kind integral equation~\eqref{} using a
Nystr\"om method.  This results in a dense linear system which we solve
with fast multipole accelerated GMRES.  Then, the shear stress,
pressure, and drag must all be computed.  Finally, the pores must be
eroded according to our erosion model~\eqref{}. \todo[inline]{Give
section numbers here}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature} 
There are several integrals that we need to compute including the
double-layer potential velocity field~\eqref{eqn:dlp}, the shear
stress~\eqref{}, and the pressure~\eqref{}.  By using carefully choosing
the quadrature rules, high-order solutions can be computed.  For all the
integrals, the trapezoid rule is used since it achieves spectral
accuracy for smooth periodic functions.  However, care must be taken
when the source point and target points coincide.

For the velocity field, the kernel of the integral operator
\begin{align}
  D(\xx,\yy) = \frac{\rr \cdot \nn}{\rho^2} 
               \frac{\rr \otimes \rr}{\rho^2}
\end{align}
has a removable singularity when $\xx = \yy$.  In particular
\begin{align}
  \lim_{\substack{\yy \rightarrow \xx \\ \yy \in \Gamma}} D(\xx,\yy) = 
  \frac{\kappa(\xx)}{2\pi}\ss \otimes \ss,
\end{align}
where $\kappa(\xx)$ is the curvature of $\Gamma$ at $\xx$.  By replacing
the diagonal term

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Time Stepping} 
We use a quasi-static formulation ....

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{$\theta-L$ Formulation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fast Summation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Artificial Diffusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Gaussian Filter}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation\label{s:formulation}} 
We consider multiple bodies being eroded by a two-dimensional Stokesian
fluid.  The Stokes paradox is circumvented by using a rounded-off
rectangle that encloses the eroding bodies.  Then, the flow is driven
by a boundary conditions, such as a Hagen-Poiseuille flow, on the
boundary of the outer geometry \todo[inline]{Need a figure here}
\begin{subequations}
\label{eqn:erosionModel}
\begin{gather}
eqn1 \label{eqn:eqn1} \\
eqn2 \label{eqn:eqn2} \\
\pderiv{\xx}{\nn} = -|\tau| \label{eqn:poreEvolution}
\end{gather}
\end{subequations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
The incompressible Stokes equations are
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Gamma.
\end{alignat*}
The viscosity $\mu$ sets the time scale, and we will assume that
$\mu=1$.  The double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\Gamma} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the unit outward
normal, solves the incompressible Stokes equations for any sufficiently
smooth density function $\eeta$.  However, the double-layer potential can not capture velocity fields
corresponding to rigid body motions~\cite{pow-mir1987}.  A standard
method to complete the double-layer potential is to introduce a
Stokeslet and rotlet due to each interior body
\begin{align}
  S[\llambda_k,\cc_k] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_k
  \quad \text{and} \quad
  R[\xi_k,\cc_k] = \frac{\rr^\perp}{\rho^2}\xi_k,
  \label{eqn:stokeslet_rotlet}
\end{align}
where $\cc_k$ is a point in the interior of $\Omega_k$, $\rr = \xx -
\cc_k$, and $\rr^\perp = (r_2,-r_1)$.  Then, the solution of the Stokes
equations with a Dirichlet boundary condition $\ff$ can be written in
terms of the completed double-layer potential
\begin{align}
  \uu(\xx) = \DD[\eeta](\xx) + \sum_{k=1}^{M} S[\llambda_k,\cc_k](\xx) +
              \sum_{k=1}^{M} R[\xi_k,\cc_k](\xx), \quad \xx \in \Omega,
  \label{eqn:completedDLP}
\end{align}
where the density function, Stokeslets, and rotlets satisfy the
second-kind integral equation
\begin{subequations}
\label{eqn:completed_DLP}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{k=1}^{M} S[\llambda_k,\cc_k](\xx) +
      \sum_{k=1}^{M} R[\xi_k,\cc_k](\xx), \quad \xx \in \Gamma, 
      \label{eqn:DLP}\\
  \llambda_k &= \frac{1}{2\pi}\int_{\Gamma_k} \eeta(\yy) \, ds_\yy, 
  \label{eqn:stokeslet} \\
  \xi_k &= \frac{1}{2\pi}\int_{\Gamma_k} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Finally, because of the incompressibility constraint, the boundary
condition must satisfy 
\begin{align}
  \int_{\Gamma} \ff \cdot \nn \, ds_\yy = 0.
\end{align}
This results in a rank-one null space of
equation~\eqref{eqn:completed_DLP} which we remove by adding the rank
one modification $\NN_0[\eeta](\xx) = \int_{\Gamma_0} (\nn(\xx) \times
\nn(\yy))\nn(\yy) \, ds_\yy$, for $\xx \in \Gamma_0$.

@@

Considering the double-layer potential~\eqref{eqn:velocityDLP} evaluated
at $\xx \in \Omega$, the deformation tensor is
\begin{align}
  \ssigma^\DD(\xx) = \int_{\bd\Omega} K_\ssigma(\xx,\yy) \eeta(\yy) \,ds_\yy,
  \label{eqn:shearStressDLP}
\end{align}
where the kernel $K_\ssigma$ is given by
\begin{align*}
  K_\ssigma(\xx,\yy) = \frac{1}{2\pi}\left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn) +
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
We are interested in the deformation tensor at boundary points $\xx_0
\in \bd\Omega$, and this requires an expression for the jump in
$\ssigma^\DD(\xx)$ as $\xx \rightarrow \xx_0 \in \Gamma$.  The limiting
value is given by~\cite{qua-bir2014a}
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}\sigma_\DD(\xx) =
  J[\eeta](\xx_0) + \ssigma^\DD(\xx_0) = 
  \frac{1}{2}\pderiv{\eeta}{\ss} \cdot \ss
  \left[ \begin{array}{cc}
    s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
  \end{array}\right] + \ssigma^\DD(\xx_0), \quad \xx_0 \in \Gamma.
\end{align}

@@

To compute the pressure at $\xx_0 \in \Gamma$, we follow the same
procedure used for the shear stress.  In particular, we compute the
pressure due to each of the three terms that compose the completed
double-layer potential.  They are given by
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}p^\DD(\xx) =  
  -\pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0),
\end{align}
where
\begin{align}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\Gamma} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta(\yy) \,ds_\yy.
\end{align}
The pressure due to the Stokeslets is
\begin{align}
  p^S(\xx_0) = \frac{\rr \cdot \llambda}{2\pi\rho^2},
  \label{eqn:pressureDLP}
\end{align}
and the velocity due to a rotlet is pressure-free.  Therefore, once the
density function and Stokeslets are computed, the pressure at a point
$\xx_0 \in \Gamma$ is
\begin{align}
  p(\xx_0) = -\pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0) + 
              \frac{\rr \cdot \llambda}{2\pi\rho^2}.
\end{align}

@@

\todo[inline]{Describe convention for normal and tangent vectors}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
We discretize the second-kind integral equation~\eqref{eqn:DLP} using a
Nystr\"om method.  The equations for the
Stokeslets~\eqref{eqn:stokeslet} and rotlets~\eqref{eqn:rotlet} are
approximated with the trapezoid rule.  By applying a quadrature method,
this results in a dense linear system which we solve with fast multipole
accelerated GMRES.  Because of the second-kind nature of the integral
equation, the number of GMRES iterations is mesh independent (but not
geometry independent), and the overall complexity of the Stokes solver
is proportional to the cost of a single matrix-vector multiplication.
The shear stress, pressure, and drag must also be computed using
quadrature, and this can be interpreted as a post-processing step once
the density function, Stokeslets, and rotlets are computed.  All the
quadrature methods are described in Section~\ref{sec:quadrature}

Once the shear stress is computed, it is used to update the shape of the
pores according to the normal flow~\eqref{eqn:poreEvolution}.  To
eliminate tangling and stretching of the mesh, we use a $\theta-L$
formulation~\cite{hou-low-she1994} which is described in
Section~\ref{sec:thetaL}.  In this section, we also describe our time
stepping method and two regularization terms that are necessary to
control the curvature of the pores and to maintain spectral accuracy in
space.

\todo[inline]{Describe the upcoming subsections better}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature} 
\label{sec:quadrature}
There are several integrals that we need to compute including the
double-layer potential velocity field~\eqref{eqn:velocityDLP}, the shear
stress~\eqref{eqn:shearStressDLP}, and the
pressure~\eqref{eqn:pressureDLP}.  By using carefully choosing the
quadrature rules, high-order solutions can be computed.  For all the
integrals, the trapezoid rule is used since it achieves spectral
accuracy for smooth periodic functions.  However, care must be taken
when the source point and target points coincide.

For the velocity field, the kernel of the integral operator
\begin{align}
  D(\xx,\yy) = \frac{\rr \cdot \nn}{\rho^2} 
               \frac{\rr \otimes \rr}{\rho^2}
\end{align}
has a removable singularity when $\xx = \yy$.  In particular
\begin{align}
  \lim_{\substack{\yy \rightarrow \xx \\ \yy \in \Gamma}} D(\xx,\yy) = 
  \frac{\kappa(\xx)}{2\pi}\ss \otimes \ss,
\end{align}
where $\kappa(\xx)$ is the curvature of $\Gamma$ at $\xx$.  By using
this limiting term in when applying the trapezoid rule, spectral
accuracy is achieved because of the smoothness and periodicity of the
integrand~\cite{tre-wei2014}.

For the shear stress and the pressure, we will use odd-even integration
which achieves spectral accuracy if the integrand is no stronger than
$\mathcal{O}(\rho^{-1})$.  Since both these integrands are too strong,
the singularity must be weakened.  This can be done by adding a constant
density function which has no effect on the shear stress and
pressure~\cite{}.  In particular, the shear stress due to the
double-layer potential is equivalent to
\begin{align}
  \ssigma^{\DD}(\xx_0) &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      \ssigma(\yy) \, ds_{\yy} \\ 
      &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      (\ssigma(\yy) - \ssigma(\xx_0)) \, ds_{\yy}.
\end{align}
For the pressure, the double-layer potential is equivalent to
\begin{align}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\Gamma} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx_0)) \,ds_\yy.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fast Summation} 
All the layer potentials are compatible with a fast summation such as
the Fast Multipole Method (FMM).  The majority of the

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{$\theta-L$ Formulation} 
\label{sec:thetaL}

We use a quasi-static approximation so that the velocity field
instantaneously changes with the changing geometry.  Therefore, we can
alternate between the two main aspects of our simulation
\begin{enumerate}
  \item Given a geometry, compute the shear stress of along each of the
  pores.
  \item Given the shear stress, update the geometry according to the
  erosion model.
\end{enumerate}

\todo[inline]{Artificial Diffusion} 
\todo[inline]{Gaussian Filter}
